---
title: "Javier Bravo Bernal"

engine: knitr
---
**LinkedIn:** [linkedin.com/in/jalfred2598](https://www.linkedin.com/in/jalfred2598)


I am a **Mechatronics Engineer** currently pursuing a Master of Science in Business Analytics (STEM). My work sits at the intersection of engineering and data, where I focus on improving systems, processes, and decision-making through structured analysis.

My experience in engineering, combined with training in SQL, Python, statistical modeling, and machine learning, allows me to approach problems with both technical depth and analytical rigor.

This portfolio highlights projects where I apply data-driven methods to solve practical, real-world challenges.

```{r}
#| echo: false
#| message: false
#| warning: false

if(!require("leaflet")){
    options(repos=c(CRAN="https://cloud.r-project.org"))
    install.packages("leaflet")
    stopifnot(require("leaflet"))
}

baruch_longitude <- -73.98333
baruch_latitude  <- +40.74028

leaflet() |>
  addTiles() |>
  setView(baruch_longitude, baruch_latitude, zoom=17) |>
  addPopups(baruch_longitude, baruch_latitude, 
            "I am a Master's student at <b>Baruch College</b>!")
```

```{r}
#| include: false
1+1
```
--------------
Last Updated: `r format(Sys.time(), "%A %m %d, %Y at %H:%M%p")`