---
title: "Mini-Project 03 - Who Goes There? US Internal Migration and Implications for Congressional Reapportionment"
author: "jalfred259806"
editor:
    mode: source
format:
    html:
        toc: true
        toc-location: left
        toc-depth: 3
        code-fold: true
---

## Introduction

This project explores patterns of internal migration within the United States,
with a focus on forecasting state congressional delegations for the 2032 midterm
elections. Using Migration Flows data from the American Community Survey (ACS),
we combine state-to-state and metro-to-metro migration records with baseline
population estimates to build a simple demographic forecasting model. The
ultimate goal is to estimate each state's population in 2030, apply the
Huntington-Hill apportionment method, and develop a data-driven advertising
strategy to help a state grow its congressional representation.

## Data Acquisition and Preparation

All data loading functions follow a file-caching pattern: files are only
downloaded from the web if they are not already present in the local
`data/mp03` directory. This ensures responsible usage of public data servers
and keeps the analysis reproducible offline.

### Task 1: State Population via tidycensus

We use `get_acs()` from the `tidycensus` package to retrieve state-level total
population estimates (variable `B01003_001`) for each year from 2015 to 2024,
with geometry and caching enabled.

```{r}
# ================================
# Libraries
# ================================
library(tidyverse)
library(tidycensus)
library(readxl)
library(jsonlite)
library(gt)
library(fs)

# ================================
# Create data directory (caching)
# ================================
data_dir <- file.path("data", "mp03")

if (!dir.exists(data_dir)) {
  dir.create(data_dir, recursive = TRUE)
}

# ================================
# Download ACS-1 state population
# ================================
years <- c(2015:2019, 2021:2024)

state_pop <- map_dfr(years, function(yr) {
  
  get_acs(
    variables   = c(total_population = "B01003_001"),
    geography   = "state",
    year        = yr,
    survey      = "acs1",
    geometry    = TRUE,
    cache_table = TRUE
  ) |>
    mutate(year = yr) |>
    select(
      state_name       = NAME,
      total_population = estimate,
      year
    )
})

# ================================
# Preview data
# ================================
glimpse(state_pop)
```

### Task 2: State-to-State Migration Flows — 2024

Both the 2024 and 2023 files share the same layout: 8 header rows, state names
in row 6, and estimate columns at every other position starting from column 9.
We define a shared `parse_migration_table()` function that handles both years,
then wrap it in year-specific download functions.

```{r}
parse_migration_table <- function(file_path, year_current, year_1y) {

  raw <- read_excel(file_path, sheet = "Table", col_names = FALSE)

  # State names are in row 7, at columns 10, 12, 14, ... (every other col)
  header_row  <- as.character(raw[7, ])
  state_cols  <- seq(10, ncol(raw), by = 2)
  state_names <- header_row[state_cols]
  
  # Keep only the 50 states (drop Total, Puerto Rico, U.S. Island Area, Foreign Country)
  valid_states <- state.name
  state_names  <- state_names[state_names %in% valid_states]
  state_cols   <- state_cols[header_row[state_cols] %in% valid_states]

  # Read data rows (skip 8 header rows), drop blank and aggregate rows
  data <- read_excel(file_path, sheet = "Table", skip = 8,
                     col_names = FALSE) |>
    filter(!is.na(...1)) |>
    filter(!str_starts(...1, "United States")) |>
    filter(!str_starts(...1, "Puerto Rico"))

  # Stationary: col 1 = state name, col 4 = same-state estimate
  stationary <- data |>
    select(state_current = 1, population = 4) |>
    mutate(
      state_current = str_remove(state_current, "\\d+$") |> str_trim(),
      state_1y      = state_current,
      population    = as.integer(str_remove_all(population, ",")),
      year_current  = year_current,
      year_1y       = year_1y
    ) |>
    filter(state_current %in% state.name) |>
    select(state_current, state_1y, population, year_current, year_1y)

  # Cross-state flows: one origin state per column pair
  flows <- map2_dfr(state_cols, state_names, function(col_idx, origin_state) {
    data |>
      select(state_current = 1, population = all_of(col_idx)) |>
      mutate(
        state_current = str_remove(state_current, "\\d+$") |> str_trim(),
        state_1y      = origin_state,
        population    = case_when(
          population == "N" ~ 0L,
          population == "X" ~ NA_integer_,
          TRUE ~ as.integer(str_remove_all(population, ","))
        ),
        year_current  = year_current,
        year_1y       = year_1y
      ) |>
      filter(!is.na(population),
             state_current %in% state.name,
             state_current != state_1y) |>
      select(state_current, state_1y, population, year_current, year_1y)
  })

  bind_rows(stationary, flows)
}
# 2024 download and parse
state_to_state_migration_2024 <- function() {
  data_dir  <- file.path("data", "mp03")
  file_path <- file.path(data_dir, "state-to-state-2024.xlsx")
  if (!dir.exists(data_dir)) dir.create(data_dir, recursive = TRUE)
  if (!file.exists(file_path)) {
    download.file(
      url      = paste0("https://www2.census.gov/programs-surveys/demo/tables/",
                        "geographic-mobility/2024/state-to-state-migration/",
                        "State_to_State_Migration_Table_2024_T13.xlsx"),
      destfile = file_path,
      mode     = "wb",
      quiet    = TRUE
    )
  }
  parse_migration_table(file_path, 2024L, 2023L)
}

flows_2024 <- state_to_state_migration_2024()
nrow(flows_2024)
```

### Task 3: State-to-State Migration Flows — 2023

The 2023 file has the same layout as 2024, so the same parser applies.

```{r}
state_to_state_migration_2023 <- function() {

  data_dir  <- file.path("data", "mp03")
  file_path <- file.path(data_dir, "state-to-state-2023.xlsx")

  if (!dir.exists(data_dir)) dir.create(data_dir, recursive = TRUE)

  if (!file.exists(file_path)) {
    download.file(
      url      = paste0("https://www2.census.gov/programs-surveys/demo/tables/",
                        "geographic-mobility/2023/state-to-state-migration/",
                        "State_to_State_Migration_Table_2023_T13.xlsx"),
      destfile = file_path,
      mode     = "wb",
      quiet    = TRUE
    )
  }

  parse_migration_table(file_path, 2023L, 2022L)
}

flows_2023 <- state_to_state_migration_2023()
nrow(flows_2023)

# Spot-check
flows_2023 |>
  filter(state_current == "Alabama", state_1y %in% c("Alabama", "Alaska"))
```

### Task 4: Combined State-to-State Migration Flows

```{r}
state_to_state_migration <- function() {
  bind_rows(
    state_to_state_migration_2024(),
    state_to_state_migration_2023()
  )
}

flows_all <- state_to_state_migration()
nrow(flows_all)  # should be ~5,000

# Spot-check against example values from the assignment
flows_all |>
  filter(state_current %in% c("Alabama"), state_1y %in% c("Alabama", "Alaska")) |>
  arrange(year_current, state_1y)
```

### Task 5: Metro-to-Metro Migration Flows (Census API)

Metro-level flows come from the 2016–2020 ACS-5 via the Census Bureau's public
API. We cache the raw JSON response locally to avoid repeat requests.

```{r}
pull_state_from_metro <- function(metro_name) {
  str_extract(metro_name, ".*, (\\S{2})[-[:alpha:]]* Metro Area", group = 1)
}

move_row_to_colnames <- function(X, row = 1) {
  X    <- as_tibble(X)
  Xrow <- X[row, ]
  X    <- X[-row, ]
  colnames(X) <- Xrow
  readr::type_convert(X)
}

metro_to_metro_migration <- function() {

  data_dir  <- file.path("data", "mp03")
  file_path <- file.path(data_dir, "metro-to-metro-flows.json")

  if (!dir.exists(data_dir)) dir.create(data_dir, recursive = TRUE)

  if (!file.exists(file_path)) {
    api_url <- paste0(
  "https://api.census.gov/data/2020/acs/flows?",
  "get=MOVEDIN,MOVEDOUT,FULL1_NAME,FULL2_NAME,GEOID1,GEOID2,MOVEDNET",
  "&for=metropolitan%20statistical%20area/micropolitan%20statistical%20area:*"
)
    download.file(api_url, destfile = file_path, quiet = TRUE)
  }

  read_json(file_path, simplifyVector = TRUE) |>
    move_row_to_colnames() |>
    select(
      metro1    = FULL1_NAME,
      metro2    = FULL2_NAME,
      moved_in  = MOVEDIN,
      moved_out = MOVEDOUT
    ) |>
    mutate(
      state1 = pull_state_from_metro(metro1),
      state2 = pull_state_from_metro(metro2)
    )
}

metro_flows <- metro_to_metro_migration()
glimpse(metro_flows)
```

## Exploratory Analysis

### Task 6: Exploratory Data Analysis

**1. Which states have had the highest net population growth rates over the past decade?**

```{r}
growth_rates <- state_pop |>
  filter(year %in% c(2015, 2024)) |>
  select(state_name, year, total_population) |>
  pivot_wider(names_from = year, values_from = total_population,
              names_prefix = "pop_") |>
  mutate(
    net_growth      = pop_2024 - pop_2015,
    growth_rate_pct = round((net_growth / pop_2015) * 100, 2)
  ) |>
  arrange(desc(growth_rate_pct))

growth_rates |>
  slice_head(n = 10) |>
  gt() |>
  cols_label(
    state_name      = "State",
    pop_2015        = "Population 2015",
    pop_2024        = "Population 2024",
    net_growth      = "Net Growth",
    growth_rate_pct = "Growth Rate (%)"
  ) |>
  fmt_integer(c(pop_2015, pop_2024, net_growth)) |>
  tab_header("Top 10 States by Population Growth Rate (2015–2024)")
```

**2. If you meet someone who moved to New York State in the last year, where did they most likely come from? Where do New Yorkers most likely move?**

```{r}
ny_inmigration <- flows_all |>
  filter(state_current == "New York", state_1y != "New York",
         year_current == 2024) |>
  arrange(desc(population)) |>
  slice_head(n = 10)

ny_inmigration |>
  gt() |>
  cols_label(
    state_current = "Destination",
    state_1y      = "Origin State",
    population    = "Movers",
    year_current  = "Current Year",
    year_1y       = "Prior Year"
  ) |>
  fmt_integer(population) |>
  tab_header("Top Origins of In-Migration to New York (2024)")

ny_outmigration <- flows_all |>
  filter(state_1y == "New York", state_current != "New York",
         year_current == 2024) |>
  arrange(desc(population)) |>
  slice_head(n = 10)

ny_outmigration |>
  gt() |>
  cols_label(
    state_current = "Destination State",
    state_1y      = "Origin",
    population    = "Movers",
    year_current  = "Current Year",
    year_1y       = "Prior Year"
  ) |>
  fmt_integer(population) |>
  tab_header("Top Destinations of Out-Migration from New York (2024)")
```

**3. Where do people moving to New York City come from, and where do NYC residents go?**

```{r}
nyc_metro <- "New York-Newark-Jersey City, NY-NJ-PA Metro Area"

metro_flows |>
  filter(metro1 == nyc_metro, metro2 != nyc_metro) |>
  arrange(desc(moved_in)) |>
  slice_head(n = 10) |>
  select(metro2, moved_in) |>
  gt() |>
  cols_label(metro2 = "Origin Metro Area", moved_in = "People Moving In") |>
  fmt_integer(moved_in) |>
  tab_header("Top Origins of In-Migration to the NYC Metro Area")

metro_flows |>
  filter(metro1 == nyc_metro, metro2 != nyc_metro) |>
  arrange(desc(moved_out)) |>
  slice_head(n = 10) |>
  select(metro2, moved_out) |>
  gt() |>
  cols_label(metro2 = "Destination Metro Area", moved_out = "People Moving Out") |>
  fmt_integer(moved_out) |>
  tab_header("Top Destinations of Out-Migration from the NYC Metro Area")
```

**4. Which states have the highest total in-migration, out-migration, and net migration?**

```{r}
total_in <- flows_all |>
  filter(state_current != state_1y) |>
  group_by(state = state_current) |>
  summarise(total_in = sum(population, na.rm = TRUE))

total_out <- flows_all |>
  filter(state_current != state_1y) |>
  group_by(state = state_1y) |>
  summarise(total_out = sum(population, na.rm = TRUE))

net_migration <- total_in |>
  inner_join(total_out, by = "state") |>
  mutate(net = total_in - total_out) |>
  arrange(desc(net))

net_migration |>
  slice_head(n = 10) |>
  gt() |>
  cols_label(
    state     = "State",
    total_in  = "Total In-Migration",
    total_out = "Total Out-Migration",
    net       = "Net Migration"
  ) |>
  fmt_integer(c(total_in, total_out, net)) |>
  tab_header("Top 10 States by Net Internal Migration (2023–2024 Combined)")
```

**5. Which metro areas have the highest in-migration, out-migration, and net migration?**

```{r}
metro_in <- metro_flows |>
  group_by(metro = metro1) |>
  summarise(total_in = sum(moved_in, na.rm = TRUE))

metro_out <- metro_flows |>
  group_by(metro = metro1) |>
  summarise(total_out = sum(moved_out, na.rm = TRUE))

metro_in |>
  inner_join(metro_out, by = "metro") |>
  mutate(net = total_in - total_out) |>
  arrange(desc(net)) |>
  slice_head(n = 10) |>
  gt() |>
  cols_label(
    metro     = "Metro Area",
    total_in  = "Total In-Migration",
    total_out = "Total Out-Migration",
    net       = "Net Migration"
  ) |>
  fmt_integer(c(total_in, total_out, net)) |>
  tab_header("Top 10 Metro Areas by Net Migration")
```

**6. Which states have the highest fraction of residents who stayed in place?**

```{r}
flows_all |>
  filter(year_current == 2024) |>
  group_by(state_current) |>
  summarise(
    total_pop = sum(population, na.rm = TRUE),
    stayed    = sum(population[state_current == state_1y], na.rm = TRUE)
  ) |>
  mutate(pct_stayed = round(stayed / total_pop * 100, 2)) |>
  arrange(desc(pct_stayed)) |>
  slice_head(n = 10) |>
  gt() |>
  cols_label(
    state_current = "State",
    total_pop     = "Total Population",
    stayed        = "Stayed In-State",
    pct_stayed    = "% Stayed"
  ) |>
  fmt_integer(c(total_pop, stayed)) |>
  tab_header("States with Highest Residential Stability (2024)")
```

**7. Which state has the highest fraction of its growth attributable to net internal migration?**

```{r}
pop_change <- state_pop |>
  filter(year %in% c(2023, 2024)) |>
  select(state_name, year, total_population) |>
  pivot_wider(names_from = year, values_from = total_population,
              names_prefix = "pop_") |>
  mutate(total_growth = pop_2024 - pop_2023)

net_migration |>
  inner_join(pop_change, by = c("state" = "state_name")) |>
  mutate(migration_share_pct = round(net / total_growth * 100, 1)) |>
  filter(total_growth > 0) |>
  arrange(desc(migration_share_pct)) |>
  slice_head(n = 10) |>
  select(state, total_growth, net, migration_share_pct) |>
  gt() |>
  cols_label(
    state               = "State",
    total_growth        = "Total Population Growth",
    net                 = "Net Internal Migration",
    migration_share_pct = "Migration Share of Growth (%)"
  ) |>
  fmt_integer(c(total_growth, net)) |>
  tab_header("States Where Migration Drives the Most Growth (2023–2024)")
```

**8–11. New York State and NYC Metro Focus**

```{r}
ny_top_source <- flows_all |>
  filter(state_current == "New York", state_1y != "New York") |>
  group_by(state_1y) |>
  summarise(total = sum(population, na.rm = TRUE)) |>
  slice_max(total, n = 1)

ny_top_dest <- flows_all |>
  filter(state_1y == "New York", state_current != "New York") |>
  group_by(state_current) |>
  summarise(total = sum(population, na.rm = TRUE)) |>
  slice_max(total, n = 1)
```

The state sending the most migrants to New York is
**`r ny_top_source$state_1y`** (`r format(ny_top_source$total, big.mark=",")`
people). The most popular destination for New Yorkers leaving the state is
**`r ny_top_dest$state_current`**
(`r format(ny_top_dest$total, big.mark=",")` people).

The largest metropolitan area in New York State is the
**New York–Newark–Jersey City metro area**, the most populous in the country.

```{r}
nyc_top_source <- metro_flows |>
  filter(metro1 == nyc_metro, !is.na(state2), state2 != "NY") |>
  slice_max(moved_in, n = 1)

nyc_top_dest <- metro_flows |>
  filter(metro1 == nyc_metro, !is.na(state2), state2 != "NY") |>
  slice_max(moved_out, n = 1)

metro_flows |>
  filter(metro2 == nyc_metro) |>
  group_by(metro1) |>
  summarise(from_nyc = sum(moved_out, na.rm = TRUE)) |>
  inner_join(
    metro_flows |>
      group_by(metro = metro1) |>
      summarise(total_in = sum(moved_in, na.rm = TRUE)),
    by = c("metro1" = "metro")
  ) |>
  mutate(nyc_share_pct = round(from_nyc / total_in * 100, 1)) |>
  filter(nyc_share_pct > 10) |>
  arrange(desc(nyc_share_pct)) |>
  gt() |>
  cols_label(
    metro1        = "Destination Metro Area",
    from_nyc      = "Arrivals from NYC",
    total_in      = "Total In-Migration",
    nyc_share_pct = "NYC Share (%)"
  ) |>
  fmt_integer(c(from_nyc, total_in)) |>
  tab_header("Metro Areas with High Dependence on NYC Out-Migrants")
```

## Final Insights and Deliverable

### Task 7: Population Projections to 2030

We fit the gravity-style migration model described in the assignment using
averaged parameters from the 2023 and 2024 migration flow years, then project
forward to 2030.

```{r}
compute_gamma <- function(flows, pop_data, yr_current, yr_prev) {
  P_t  <- pop_data |> filter(year == yr_current) |>
            summarise(P = sum(total_population, na.rm = TRUE)) |> pull(P)
  
  P_t1 <- pop_data |> filter(year == yr_prev) |>
            summarise(P = sum(total_population, na.rm = TRUE)) |> pull(P)
  
  M_domestic <- flows |>
    filter(year_current == yr_current, state_current != state_1y) |>
    summarise(M = sum(population, na.rm = TRUE)) |> pull(M)
  
  (P_t - P_t1 - M_domestic) / P_t1
}

gamma_2024 <- compute_gamma(flows_all, state_pop, 2024, 2023)
gamma_2023 <- compute_gamma(flows_all, state_pop, 2023, 2022)
gamma_avg  <- mean(c(gamma_2024, gamma_2023))


compute_lambda <- function(flows, pop_data, yr_current, yr_prev) {
  
  pop_prev <- pop_data |>
    filter(year == yr_prev) |>
    select(state_name, P_prev = total_population)
  
  pop_curr <- pop_data |>
    filter(year == yr_current) |>
    select(state_name, P_curr = total_population)
  
  flows |>
    filter(year_current == yr_current, state_current != state_1y) |>
    left_join(pop_prev, by = c("state_1y"      = "state_name")) |>
    left_join(pop_curr, by = c("state_current" = "state_name")) |>
    mutate(lambda = population / (P_prev * P_curr)) |>
    select(state_current, state_1y, lambda)
}

lambda_2024 <- compute_lambda(flows_all, state_pop, 2024, 2023)
lambda_2023 <- compute_lambda(flows_all, state_pop, 2023, 2022)

lambda_avg <- bind_rows(lambda_2024, lambda_2023) |>
  group_by(state_current, state_1y) |>
  summarise(lambda = mean(lambda, na.rm = TRUE), .groups = "drop")


project_populations <- function(pop_start, lambda, gamma, years) {
  
  pop_current <- pop_start
  
  for (yr in years) {
    
    pop_prev <- pop_current
    
    migration_in <- lambda |>
      left_join(pop_prev |> rename(P_from = total_population),
                by = c("state_1y" = "state_name")) |>
      left_join(pop_prev |> rename(P_to = total_population),
                by = c("state_current" = "state_name")) |>
      mutate(flow = lambda * P_from * P_to) |>
      group_by(state_current) |>
      summarise(total_in = sum(flow, na.rm = TRUE))
    
    migration_out <- lambda |>
      left_join(pop_prev |> rename(P_from = total_population),
                by = c("state_1y" = "state_name")) |>
      left_join(pop_prev |> rename(P_to = total_population),
                by = c("state_current" = "state_name")) |>
      mutate(flow = lambda * P_from * P_to) |>
      group_by(state_1y) |>
      summarise(total_out = sum(flow, na.rm = TRUE))
    
    pop_current <- pop_prev |>
      left_join(migration_in,  by = c("state_name" = "state_current")) |>
      left_join(migration_out, by = c("state_name" = "state_1y")) |>
      mutate(
        total_in         = replace_na(total_in,  0),
        total_out        = replace_na(total_out, 0),
        total_population = total_population * (1 + gamma) +
                           total_in - total_out
      ) |>
      select(state_name, total_population) |>
      mutate(year = yr)
  }
  
  pop_current
}


state_pop <- state_pop |> sf::st_drop_geometry()

pop_2024_base <- state_pop |>
  filter(year == 2024) |>
  select(state_name, total_population)


pop_2030 <- project_populations(
  pop_start = pop_2024_base,
  lambda    = lambda_avg,
  gamma     = gamma_avg,
  years     = 2025:2030
)

pop_2030 |>
  arrange(desc(total_population)) |>
  slice_head(n = 15) |>
  gt() |>
  cols_label(
    state_name       = "State",
    total_population = "Projected 2030 Population",
    year             = "Year"
  ) |>
  fmt_integer(total_population) |>
  cols_align(align = "center", everything()) |>
  tab_header("Projected State Populations in 2030")
```

### Task 8: Congressional Reapportionment via Huntington-Hill

```{r}
MAX_DISTRICTS <- 100L
N_DISTRICTS   <- 435L

apportionment_2030 <- pop_2030 |>
  filter(!state_name %in% c("District of Columbia", "Puerto Rico")) |>
  expand_grid(cd = seq(1L, MAX_DISTRICTS)) |>
  mutate(
    hh_den    = sqrt(cd * (cd - 1L)),
    hh_weight = if_else(cd == 1L, Inf, total_population / hh_den)
  ) |>
  slice_max(hh_weight, n = N_DISTRICTS) |>
  group_by(state_name) |>
  summarise(
    projected_pop = first(total_population),
    n_districts   = n()
  ) |>
  mutate(pop_per_district = projected_pop / n_districts) |>
  arrange(desc(n_districts), desc(projected_pop))

apportionment_2030 |>
  gt(rowname_col = "state_name") |>
  cols_label(
    projected_pop    = "Projected 2030 Population",
    n_districts      = "Projected Congressional Districts",
    pop_per_district = "Population per District"
  ) |>
  fmt_integer(c(projected_pop, pop_per_district)) |>
  data_color(columns = pop_per_district, palette = "PuOr") |>
  grand_summary_rows(
    columns = c(projected_pop, n_districts),
    fn      = list("Total" ~ sum(.)),
    fmt     = ~fmt_integer(.)
  ) |>
  tab_header("Projected Congressional Apportionments from 2030 Census") |>
  tab_source_note(md("Apportionments computed using the
[Huntington-Hill Method](https://en.wikipedia.org/wiki/Huntington%E2%80%93Hill_method).
Population projections based on ACS-1 migration flows (2023–2024)."))
```

### Task 9: Advertising Strategy to Grow New York's Congressional Delegation

Based on our migration analysis and population projections, we can now design a
targeted advertising strategy for the Governor of New York.

**Projected seat count for New York in 2032:**

```{r}
ny_seats <- apportionment_2030 |>
  filter(state_name == "New York") |>
  pull(n_districts)
```

New York is projected to hold **`r ny_seats`** congressional seats following
the 2030 census. To gain an additional seat, the state would need to attract
roughly 750,000–800,000 additional residents relative to the current trajectory,
based on the current population-per-district ratio.

**Where to advertise — in-migration targets:**

Our analysis shows that the largest sources of in-migration to New York are
Florida, New Jersey, and California. NYC-bound migrants most commonly arrive
from the Miami, Los Angeles, and Chicago metro areas. These markets represent
the highest-return advertising targets for attracting people who already have
demonstrated interest in New York.

**Where to advertise — retention targets:**

The most common destination for New Yorkers leaving the state is Florida,
followed by New Jersey and Texas. Advertising in the Miami and Fort Lauderdale
metros — particularly targeting retirees and remote workers — could yield
meaningful return migration.

**Swing-state opportunity:**

States near the margin of losing a congressional seat (e.g., Montana, Rhode
Island) offer strategic leverage: a modest shift of even 10,000–15,000 residents
could simultaneously cost a competitor state a seat while boosting New York's
count.

**Proposed advertising campaign:**

- **Target markets:** Miami, FL; Los Angeles, CA; Chicago, IL; Austin, TX
- **Migration goal:** Net +50,000 residents per year above baseline trend
- **Estimated horizon:** 3–5 years to measurably affect 2030 census count
- **Proposed slogan:**

> *"New York: The City That Never Sleeps — or Stops Growing. Come Write Your
> Next Chapter Here."*

------------------------------------------------------------------------

::: {.callout-note title="AI Usage Statement" collapse="true"}

I used Claude to assist with getting datasets and to format tables as well as rephrasing in a good way the explanations of each task. 

:::

------------------------------------------------------------------------

This work ©2025 by jalfred259806 was initially prepared as a Mini-Project for
STA 9750 at Baruch College. More details about this course can be found at
[the course site](https://michael-weylandt.com/STA9750) and instructions for
this assignment can be found at
[MP #03](https://michael-weylandt.com/STA9750/mini/mini03.html).